---
title: Visualise a Multiple Linear Regression (MLR) Model
Source: UC Applied Data Science
May 2025
---

Solve the coefficient vector for a Multiple Linear Regression (MLR) model using the Auto dataset. Set up the linear algebra, compute the coefficients analytically, and visualise the 3 variables graphically. Finalise generalising the model to a third feature without visualisation.

```{r}
#install.packages("ISLR")
#install.packages(c("plotly", "dplyr"))  # Install plotly and dplyr (for %>%)
library(ISLR)
library(plotly)
library(dplyr)
```

```{r}

#load the data
data("Auto")
str(Auto)
```

```{r}
#Create a training and test split. 
set.seed(123)
indices <- sample(1:nrow(Auto), 0.8*nrow(Auto))
train <- Auto[indices,]
test <- Auto[-indices,]
```

We will use `weight` and `cylinders` as predictors and `mpg` as response. Some matrix operations are involved in this formula. You can use `solve()` to get the inverse of a matrix, `t()` to transpose a matrix and `%*%` to do matrix multiplication.

```{r}
# Create an intercept vector. When multiplied by its corresponding coefficient in the regression equation, it contributes the intercept value to the model. 
intercept <- rep(1, length(indices))

# Create the design matrix A. 
weight <- train$weight
cylinders <- train$cylinders
mpg <- train$mpg
A <- as.matrix(cbind(intercept, weight, cylinders))

# Prep y for matrix operation. 
y <- as.matrix(mpg)

# Can you solve b using the analytical solution formula and the functions introduced? 
b <- solve(t(A) %*% A) %*% t(A) %*% y #
b
```

We can visualise the model defined by the coefficient vector as a 2d plane.

```{r}
# Create a grid of x1 and x2 values for plotting the regression plane
weight_grid <- seq(min(weight), max(weight), length.out = 30)
cylinders_grid <- seq(min(cylinders), max(cylinders), length.out = 30)
grid <- expand.grid(weight = weight_grid, cylinders = cylinders_grid)

# Calculate the corresponding y values on the plane
plane <- b[1] + b[2] * grid$weight + b[3] * grid$cylinders

# Plot the points and the regression plane
plot_ly() %>%
  add_markers(x = weight, y = cylinders, z = mpg, marker = list(color = 'blue', size = 4)) %>%
  add_surface(x = weight_grid, y = cylinders_grid, z = matrix(plane, nrow = 30, ncol = 30), opacity = 0.6, colorscale = list(c(0, 1), c("red", "red")), showscale = FALSE, name = 'Regression Plane') %>%
  layout(scene = list(
    xaxis = list(title = 'weight'),
    yaxis = list(title = 'cylinders'),
    zaxis = list(title = 'mpg')
  ), showlegend = FALSE)
```

We can also include another feature, `acceleration` for instance, in fitting a multiple linear regression model in the same way as before. However, it will become tricky to visualise the model.

```{r}
acceleration <- train$acceleration
new_A <- as.matrix(cbind(A,acceleration))
new_b <- solve(t(new_A) %*% new_A) %*% t(new_A) %*% y
new_b
```

### Summary

-   Coefficients for the 2-feature and 3-feature models.

-   A 3D plot of the regression plane (for 3 variables).

-   Understanding of how MLR works beyond simple linear regression.
